---
title: "exNGT: an overview of various correlation and other analyses"
author: "Thomas Münch"
date: "`r Sys.Date()`"
output:
  rmarkdown::pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(
  root.dir = "../"
)
```

## Introduction

This document summarises various analyses of the NGT firn core stack:

* comparison of the NGT stack with other temperature data sets from the Arctic2k
  compilation and weather station measurements;
* comparison of individual NGT firn cores with the Arctic2k data set;
* comparison of the mean temperature anomalies derived from the NGT stack
  between the recent 15 year period and the reference period 1961-1990 as well
  as the entire 20th century.

## Prepare data

Load relevant source files and packages:  
```{r, echo = TRUE, results = "hide", message = FALSE}
source("init.R")
```

Set the running mean filter window size (yr) and the isotope--temperature slope
(K/$\text{\textperthousand}$):
```{r}
filter.window <- 11
permil2temperature <- 1 / 0.67
```

Load the NGT data, filter it and calculate the isotope stack:
```{r}
stackedNGT <- processNGT() %>%
  stackNGT()

filteredStackedNGT <- processNGT() %>%
  filterData(window = filter.window) %>%
  stackNGT()
```

Load the Arctic2k, weather station, MAR and GBI data and filter it:
```{r}
Arctic2k <- readArctic2k() %>%
  extendWithHadCrut() %>%
  dplyr::select("Year", "TempAnomaly")

filteredExtendedArctic2k <- Arctic2k %>%
  filterData(window = filter.window)

filteredArctic2k <- readArctic2k() %>%
  filterData(window = filter.window)

filteredDMI <- readDMI() %>%
  makeAnomalies() %>%
  filterData(window = filter.window)

filteredMAR <- readMAR() %>%
  filterData(window = filter.window)

filteredGBI <- readGBI() %>%
  filterData(window = filter.window)
```

Convert the NGT stack data to temperature:
```{r}
stackedTemperatureNGT <- stackedNGT %>%
  dplyr::mutate(stack = permil2temperature * stack)

filteredStackedTemperatureNGT <- filteredStackedNGT %>%
  dplyr::mutate(stack = permil2temperature * stack)
```

## Comparison of NGT-2012 temperature stack with Arctic2k

* Overall correlation between NGT-2012 temperature stack and Arctic2k
  compilation from 1000 CE onwards with and without including HadCrut extension:
  ```{r}
  analysis.period <- 2011 : 1000
  res <- estimateCorrelation(stackedTemperatureNGT,
                             filteredStackedTemperatureNGT, filteredExtendedArctic2k,
                             filter.window = filter.window,
                             analysis.period = analysis.period,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.4f)", res$r, res$p)

  analysis.period <- 2000 : 1000
  res <- estimateCorrelation(stackedTemperatureNGT,
                             filteredStackedTemperatureNGT, filteredArctic2k,
                             filter.window = filter.window,
                             analysis.period = analysis.period,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.4f)", res$r, res$p)
  ```

* Correlation between NGT temperature stack and Arctic2k compilation for the
  20th century and for the 20th century including up to 2011 CE:
  ```{r}
  analysis.period <- 2000 : 1901
  res <- estimateCorrelation(stackedTemperatureNGT,
                             filteredStackedTemperatureNGT, filteredArctic2k,
                             filter.window = filter.window,
                             analysis.period = analysis.period,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)

  analysis.period <- 2011 : 1901
  res <- estimateCorrelation(stackedTemperatureNGT,
                             filteredStackedTemperatureNGT, filteredExtendedArctic2k,
                             filter.window = filter.window,
                             analysis.period = analysis.period,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)
  ```

* Linear trends (in degree Celsius / 100 years) of Arctic2k and NGT-2012 stack
  over pre-industrial and industrial era:
  ```{r}
  t1 <- 1800 : 1000
  t2 <- 2011 : 1800

  regressionData <- list(
    data.frame(x = t1, y = subsetData(stackedTemperatureNGT, t1, "stack")),
    data.frame(x = t2, y = subsetData(stackedTemperatureNGT, t2, "stack")),
    data.frame(x = t1, y = subsetData(Arctic2k, t1, "TempAnomaly")),
    data.frame(x = t2, y = subsetData(Arctic2k, t2, "TempAnomaly"))
  )

  regressionSlopes <- regressionData %>%
    sapply(function(df) {unname(coef(summary(lm(y ~ x, df)))[2, 1 : 2])})
  ```

  ```{r}
  # NGT 1000-1800 CE:
  (regressionSlopes[1, 1] * 100) %>% round(digits = 2)
  # +/-
  (regressionSlopes[2, 1] * 100) %>% round(digits = 2)

  # Arctic2k 1000-1800 CE:
  (regressionSlopes[1, 3] * 100) %>% round(digits = 2)
  # +/-
  (regressionSlopes[2, 3] * 100) %>% round(digits = 2)
  
  # NGT 1800-2011 CE:
  (regressionSlopes[1, 2] * 100) %>% round(digits = 2)
  # +/-
  (regressionSlopes[2, 2] * 100) %>% round(digits = 2)

  # Arctic2k 1800-2011 CE:
  (regressionSlopes[1, 4] * 100) %>% round(digits = 2)
  # +/-
  (regressionSlopes[2, 4] * 100) %>% round(digits = 2)
  ```

## Comparison of individual NGT firn core records with Arctic2k

We here use those NGT records which are also part of the Arctic2k compilation to
assess the extent to which these records contribute to the overall Arctic2k
temperature reconstruction. These firn cores are GRIP, GISP2, NGRIP, B16, B18
and B21 and we use the merged versions when redrilled cores are available.

```{r}
mergedFilteredNGT <- processNGT() %>%
  filterData(window = filter.window) %>%
  stackNGT(stack = FALSE)

t <- 2000 : 1000
a2k <- subsetData(filteredArctic2k, t, "TempAnomaly")
grip <- subsetData(mergedFilteredNGT, t, "GRIP")
gisp2 <- subsetData(mergedFilteredNGT, t, "GISP2")
ngrip <- subsetData(mergedFilteredNGT, t, "NGRIP")
b16 <- subsetData(mergedFilteredNGT, t, "B16")
b18 <- subsetData(mergedFilteredNGT, t, "B18")
b21 <- subsetData(mergedFilteredNGT, t, "B21")
```

```{r}
cor(grip, a2k, use = "pairwise") %>%
  round(digits = 2)

cor(gisp2, a2k, use = "pairwise") %>%
  round(digits = 2)

cor(ngrip, a2k, use = "pairwise") %>%
  round(digits = 2)

cor(b16, a2k, use = "pairwise") %>%
  round(digits = 2)

cor(b18, a2k, use = "pairwise") %>%
  round(digits = 2)

cor(b21, a2k, use = "pairwise") %>%
  round(digits = 2)
```


## Comparison of NGT isotope stack with DMI temperature data

We calculate the correlations of the NGT isotope stack with the temperature data
from three coastal Greenland weather stations for the time period 1901--2011.
```{r}
pituffik <- dplyr::select(filteredDMI, "Year", "Pituffik")
upernavik <- dplyr::select(filteredDMI, "Year", "Upernavik")
danmarkshavn <- dplyr::select(filteredDMI, "Year", "Danmarkshavn")
```

* Correlation of NGT isotope stack and Pituffik weather station temperature:
  ```{r}
  res <- estimateCorrelation(stackedNGT, filteredStackedNGT, pituffik,
                             filter.window = filter.window,
                             analysis.period = 2011 : 1901,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)
  ```

* Correlation of NGT isotope stack and Upernavik weather station temperature:
  ```{r}
  res <- estimateCorrelation(stackedNGT, filteredStackedNGT, upernavik,
                             filter.window = filter.window,
                             analysis.period = 2011 : 1901,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)
```

* Correlation of NGT isotope stack and Danmarkshavn weather station temperature:
  ```{r}
  res <- estimateCorrelation(stackedNGT, filteredStackedNGT, danmarkshavn,
                             filter.window = filter.window,
                             analysis.period = 2011 : 1901,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)
```

## Comparison of NGT-2012 isotope stack with MAR3.5 and GBI data

We calculate the correlation of the NGT isotope stack with data from the
regional climate model MAR version 3.5; specifically, with the annual 2m surface
air temperature extracted and averaged over the NGT-2012 firn core locations, and
with the Greenland-wide melt runoff rate, using 11-yr running mean time series
over the common period from 1871 to 2011 CE in both cases. We also analyse how
much weaker the correlation is for annual mean data.

* Correlation with MAR temperature:
```{r}

  # filtered data
  res <- estimateCorrelation(stackedNGT, filteredStackedNGT,
                             filteredMAR[, c("Year", "t2m")],
                             filter.window = filter.window,
                             analysis.period = 2011 : 1871,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)

  # annual data
  mar <- readMAR()
  res <- estimateCorrelation(stackedNGT, stackedNGT,
                             mar[, c("Year", "t2m")], filter.window = 1,
                             analysis.period = 2011 : 1871, nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)
```
* Correlation with MAR melt runoff:
```{r}
  # filtered data
  res <- estimateCorrelation(stackedNGT, filteredStackedNGT,
                             filteredMAR[, c("Year", "melt")],
                             filter.window = filter.window,
                             analysis.period = 2011 : 1871,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)
  # annual data
  res <- estimateCorrelation(stackedNGT, stackedNGT,
                             mar[, c("Year", "melt")], filter.window = 1,
                             analysis.period = 2011 : 1871, nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)

```

Additionally, we correlate both the NGT isotope data and the MAR melt runoff
rate with the Greenland Blocking Index (GBI), where we compare the correlation
for melt runoff between annual mean GBI and summer GBI values.

* Correlation between NGT and annual mean GBI:
```{r}

  # filtered data
  res <- estimateCorrelation(stackedNGT, filteredStackedNGT,
                             filteredGBI[, c("Year", "annual")],
                             filter.window = filter.window,
                             analysis.period = 2011 : 1851,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)

  # annual data
  gbi <- readGBI()
  res <- estimateCorrelation(stackedNGT, stackedNGT,
                             gbi[, c("Year", "annual")], filter.window = 1,
                             analysis.period = 2011 : 1851, nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)
```
* Correlation between MAR melt runoff and annual mean GBI:
```{r}

  # filtered data
  res <- estimateCorrelation(mar[, c("Year", "melt")],
                             filteredMAR[, c("Year", "melt")],
                             filteredGBI[, c("Year", "annual")],
                             filter.window = filter.window,
                             analysis.period = 2011 : 1871,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)

  # annual data
  res <- estimateCorrelation(mar[, c("Year", "melt")], mar[, c("Year", "melt")],
                             gbi[, c("Year", "annual")], filter.window = 1,
                             analysis.period = 2011 : 1871, nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)
```
* Correlation between MAR melt runoff and summer GBI:
```{r}

  # filtered data
  res <- estimateCorrelation(mar[, c("Year", "melt")],
                             filteredMAR[, c("Year", "melt")],
                             filteredGBI[, c("Year", "summer")],
                             filter.window = filter.window,
                             analysis.period = 2011 : 1871,
                             nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)

  # annual data
  res <- estimateCorrelation(mar[, c("Year", "melt")], mar[, c("Year", "melt")],
                             gbi[, c("Year", "summer")], filter.window = 1,
                             analysis.period = 2011 : 1871, nmc = 10000)
  sprintf("r = %1.2f (p = %1.3f)", res$r, res$p)
```

## Temperature increase of recent 11 years relative to past

We compare the mean temperature over the recent 11 years of the NGT stack with
the mean value of the 1961-1990 period and the mean value of the 20th century
(1901--2000), including an estimation of the standard error of the
difference. We do this assuming the standard spatial isotope--temperature
calibration slope for Greenland as well as assuming stronger and weaker values
from the temporal slopes derived by Vinther et al. (2009) and Masson-Delmotte et
al. (2015).

Set the alternative isotope--temperature calibration slopes:
```{r}
permil2temperature.max <- 2.1     # Vinther et al. (2009)
permil2temperature.min <- 1 / 1.1 # Masson-Delmotte et al. (2015)
```

* Recent 11 years versus 1961-1990 reference period:
  ```{r}
  t1 <- 2011 : 2001
  t2 <- 1990 : 1961

  recentPeriod <- stackedNGT %>%
    subsetData(t1, var = "stack") %>%
    dplyr::tibble() %>%
    setNames("d18O") %>%
    dplyr::summarise(mean1 = mean(d18O), se1 = sdError(d18O))
  referencePeriod <- stackedNGT %>%
    subsetData(t2, var = "stack") %>%
    dplyr::tibble() %>%
    setNames("d18O") %>%
    dplyr::summarise(mean2 = mean(d18O), se2 = sdError(d18O))

  difference <- cbind(recentPeriod, referencePeriod)

  # mean difference and s.e. for spatial slope
  (difference * permil2temperature) %>%
    dplyr::transmute(deltaT = mean1 - mean2, se = sqrt(se1^2 + se2^2)) %>%
    round(digits = 1)

  # mean difference and s.e. for minimum slope
  (difference * permil2temperature.min) %>%
    dplyr::transmute(deltaT = mean1 - mean2, se = sqrt(se1^2 + se2^2)) %>%
    round(digits = 1)

  # mean difference and s.e. for maximum slope
  (difference * permil2temperature.max) %>%
    dplyr::transmute(deltaT = mean1 - mean2, se = sqrt(se1^2 + se2^2)) %>%
    round(digits = 1)

  ```

* Recent 11 years versus 20th century (1901-2000):
  ```{r}
  t2 <- 2000 : 1901

  referencePeriod <- stackedNGT %>%
    subsetData(t2, var = "stack") %>%
    dplyr::tibble() %>%
    setNames("d18O") %>%
    dplyr::summarise(mean2 = mean(d18O), se2 = sdError(d18O))

  difference <- cbind(recentPeriod, referencePeriod)

  # mean difference and s.e. for spatial slope
  (difference * permil2temperature) %>%
    dplyr::transmute(deltaT = mean1 - mean2, se = sqrt(se1^2 + se2^2)) %>%
    round(digits = 1)

  # mean difference and s.e. for minimum slope
  (difference * permil2temperature.min) %>%
    dplyr::transmute(deltaT = mean1 - mean2, se = sqrt(se1^2 + se2^2)) %>%
    round(digits = 1)

  # mean difference and s.e. for maximum slope
  (difference * permil2temperature.max) %>%
    dplyr::transmute(deltaT = mean1 - mean2, se = sqrt(se1^2 + se2^2)) %>%
    round(digits = 1)

  ```

## References

Masson-Delmotte, V., Steen-Larsen, H. C., Ortega, P., Swingedouw, D., Popp, T.,
Vinther, B. M., Oerter, H., Sveinbjornsdottir, A. E., Gudlaugsdottir, H., Box,
J. E., Falourd, S., Fettweis, X., Gallée, H., Garnier, E., Gkinis, V., Jouzel,
J., Landais, A., Minster, B., Paradis, N., Orsi, A., Risi, C., Werner, M. and
White, J. W. C., Recent changes in north-west Greenland climate documented by
NEEM shallow ice core data and simulations, and implications for
past-temperature reconstructions, The Cryosphere, 9(4), 1481--1504,
https://doi.org/10.5194/tc-9-1481-2015, 2015.
  
Vinther, B. M., Buchardt, S. L., Clausen, H. B., Dahl-Jensen, D., Johnsen,
  S. J., Fisher, D. A., Koerner, R. M., Raynaud, D., Lipenkov, V., Andersen,
  K. K., Blunier, T., Rasmussen, S. O., Steffensen, J. P. and Svensson, A. M.,
  Holocene thinning of the Greenland ice sheet, Nature, 461(7262), 385--388,
  https://doi.org/10.1038/nature08355, 2009.

