---
title: "exNGT: summary of probabilities for the recent value to occur under the preindustrial distribution"
author: "Thomas Münch"
date: "`r Sys.Date()`"
output:
  rmarkdown::pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(
  root.dir = "../"
)
```


```{r, echo = FALSE, results = "hide", message = FALSE}

source("init.R")

# vary the stacking method
method <- c("main", "main", "main", "simple", "simple", "main")
variant <- c("", "forward diffused", "frozen (N = 5)", "", "forward diffused",
             "non-adjusted mean")
pi <- rep("default", 6)
rec <- rep("default", 6)
window <- rep("default", 6)

p <- sprintf("%1.2e", c(

  plotHistogram(stack.method = "main", plot = FALSE),

  plotHistogram(stack.method = "main", diffuse = TRUE, plot = FALSE),

  plotHistogram(stack.method = "fix_N", nfix = 5, plot = FALSE),

  plotHistogram(stack.method = "stack_all", plot = FALSE),

  plotHistogram(stack.method = "stack_all", diffuse = TRUE, plot = FALSE),

  plotHistogram(stack.method = "main", adjustMean = FALSE, plot = FALSE)
))

df1 <- data.frame(method, variant, pi, rec, window, p)

# vary the filter window
method <- rep("main", 6)
variant <- rep("", 6)
pi <- rep("default", 6)
rec <- rep("default", 6)
window <- c(3, 5, 7, 9, 15, 21)

p <- sprintf("%1.2e", c(

  plotHistogram(stack.method = "main", filter.window = window[1], plot = FALSE),

  plotHistogram(stack.method = "main", filter.window = window[2], plot = FALSE),

  plotHistogram(stack.method = "main", filter.window = window[3], plot = FALSE),

  plotHistogram(stack.method = "main", filter.window = window[4], plot = FALSE),

  plotHistogram(stack.method = "main", filter.window = window[5], plot = FALSE),

  plotHistogram(stack.method = "main", filter.window = window[6], plot = FALSE)
))

df2 <- data.frame(method, variant, pi, rec, window = as.factor(window), p)

# vary the pI period
method <- rep("main", 3)
variant <- c("", "merge at start", "merge at end")
pi <- c("1000 - 1850", "1000 - 1900", "1000 - 1900")
rec <- rep("default", 3)
window <- rep("default", 3)

p <- sprintf("%1.2e", c(

  plotHistogram(stack.method = "main", piPeriod = 1000 : 1850,
                          mergePoint = "start", plot = FALSE),

  plotHistogram(stack.method = "main", piPeriod = 1000 : 1900,
                          mergePoint = "start", plot = FALSE),

  plotHistogram(stack.method = "main", piPeriod = 1000 : 1900,
                          mergePoint = "end", plot = FALSE)
))

df3 <- data.frame(method, variant, pi, rec, window, p)

# vary the recent period
method <- rep("main", 5)
variant <- rep("", 5)
pi <- rep("default", 5)
rec <- c("2000 - 2010", "1999 - 2009", "1998 - 2008", "1997 - 2007", "1996 - 2006")
window <- rep("default", 5)

p <- sprintf("%1.2e", c(

  plotHistogram(stack.method = "main", endRecentPeriod = 2010, plot = FALSE),

  plotHistogram(stack.method = "main", endRecentPeriod = 2009, plot = FALSE),

  plotHistogram(stack.method = "main", endRecentPeriod = 2008, plot = FALSE),

  plotHistogram(stack.method = "main", endRecentPeriod = 2007, plot = FALSE),

  plotHistogram(stack.method = "main", endRecentPeriod = 2006, plot = FALSE)  
))

df4 <- data.frame(method, variant, pi, rec, window, p)



df <- rbind(df1, df2, df3, df4)

```
The following table lists probabilities for the recent NGT-2012 value to occur
under the pre-indistrial distribution depending on different parameter
combinations underlying the analysis. The source code for computing the
probabilities is hidden here due to its length and can be viewed in the R
markdown file corresponding to this document. The first five analysis variations
listed in the table correspond to the histograms shown in the figure
"supplement-histograms.pdf".

```{r, echo = FALSE}
caption <- paste("Probability for the recent value to occur under the",
             "pre-industrial distribution for different parameter combinations",
             "of creating and analysing the NGT-2012 stack. The meaning of the",
             "settings named \"default\" are: Pre-industrial period = ",
             "1000 - 1800 CE, Recent period = 2001 - 2011 CE,",
             "Filter window = 11 yr.")

knitr::kable(df, col.names = c("Method", "Variant", "Pre-industrial period",
                               "Recent period", "Filter window", "Probability"),
             caption = caption)
```
